---
title: "R-Assignment"
output: html_document
date: "2025-03-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
setwd("/Users/jordyn/Desktop/BCB546_Spring2025/assignments/Data files")
library(dplyr)
```

#### assigning txt files to fang or snp

```{r}
fang <- fang_et_al_genotypes
snp <- snp_position

dim(fang) # dim 2783, 986. this lists colms and rows
dim(snp) # dim 983, 15. this lists colms and rows

ncol(fang) # number of columns is 986
ncol(snp) # number of columns is 15

nrow(fang) # 2783 rows
nrow(snp) # 983 rows

names(snp) # tells the names of column labels

head(fang, 10) # shows first 10 rows of data
head(snp, 10) # shows first 10 rows of data

tail(fang, 10) # shows last 10 rows of data
tail(snp, 10) # shows last 10 rows of data
```

#### this forloop displays the contents of each variable in the snp file

```{r}

for(i in 1:ncol(snp)) {
  print(names(snp)[i])
  print(table(snp[, i], useNA = "ifany"))
  cat("\n") # Adds a line before the next iteration
} 
```

#### same thing but for fang file

```{r}
for(i in 1:ncol(fang)) {
  print(names(fang)[i])
  print(table(fang[, i], useNA = "ifany"))
  cat("\n") # Adds a line before the next iteration
} 
```

#### this shows the length, class, mode, min's/maxes, quantiles of the file

```{r}
summary(snp)
```

## Data Processing

#### read the file

```{r}
genotype_data <- read.table(file = "fang_et_al_genotypes.txt", header = TRUE)
```

#### For Maize: filter these specified groups, ZMMIL, ZMMLR, ZMMR and assign to genotypes_maize. Essentially, this is extracting the rows when the group is one of the three indicated groups

```{r}
genotypes_maize <- filter(genotype_data, Group == "ZMMIL" | Group == "ZMMLR" | Group == "ZMMR")

print(genotypes_maize)
```

#### Same thing but for teosinte data

```{r}
genotypes_teosinte <- filter(genotype_data, Group == "ZMPBA" | Group == "ZMPIL" | Group == "ZMPJA")

print(genotypes_teosinte)
```

#### Transpose the data for maize and teosinte

```{r}
maize_transpose <- t(genotypes_maize)
teosinte_transpose <- t(genotypes_teosinte)

print(maize_transpose)
print(teosinte_transpose)
```

#### remove the unwanted columns from the snp data as indicated by the snp[-c(2,5:15)]. This means to remove column 2 and 5 through 15.

```{r}
SNP_clean <- snp[-c(2,5:15)]
SNP_clean # we can see we have the columns we want: SNP_ID, Chromosome, and Position
```

#### label rownames as SNP_ID so we can then merge SNP_clean and the teosinte and maize files by row.names

```{r}
rownames(SNP_clean) <- SNP_clean$SNP_ID
rownames(SNP_clean)
```

#### merge maize and snp_clean

```{r}
joined_maize <- merge(SNP_clean, maize_transpose, by = "row.names")
```

#### merge teosinte and snp_clean

```{r}
joined_teosinte <- merge(SNP_clean, teosinte_transpose, by = "row.names")
```

#### create a for-loop that cycles through chromosomes 1-10 in the joined_maize data file in my global env

```{r}
for (i in 1:10) {chr_data <- subset(joined_maize, joined_maize[[3]] == i)  # Filters rows where the third column matches the chromosome number
chr_data[is.na(chr_data)] <- "?" # Replaces NA values with "?"
chr_data <- chr_data[order(chr_data[[2]]), ]  # Sorts the data by the second column in increasing order
assign(paste0("/Users/jordyn/Desktop/BCB546_Spring2025/assignments/R-Assignment/chr", i, "_increasing_maize"), chr_data)
}  # Saves the result as a new dataframe in the environment
```

#### same thing but for teosinte data

```{r}
for (i in 1:10) {chr_teosinte <- subset(joined_teosinte, joined_teosinte[[3]] == i) # Creates a new variable name for each chromosome dataset
chr_teosinte[is.na(chr_teosinte)] <- "?"  # Replaces NA values with "?"
chr_teosinte <- chr_teosinte[order(chr_teosinte[[2]]), ]  # Sorts the data by the second column in increasing order
assign(paste0("/Users/jordyn/Desktop/BCB546_Spring2025/assignments/R-Assignment/chr", i, "_increasing_teosinte"), chr_teosinte)
}   # Saves as a separate dataframe in the environment
```

#### Maize in decreasing order replacing "?" with "-"

```{r}
for (i in 1:10) {
  chr_maize <- subset(joined_maize, joined_maize[[3]] == i) # Creates a new variable for each chromosome dataset
  chr_maize[chr_maize == "?"] <- "-" # Replaces "?" with "-"
  chr_maize <- chr_maize[order(chr_maize[[2]], decreasing = TRUE), ] # Sorts by the second column in decreasing order
  assign(paste0("/Users/jordyn/Desktop/BCB546_Spring2025/assignments/R-Assignment/chr", i, "_decreasing_maize"), chr_maize) # Saves as a separate dataframe in the environment
  write.table(chr_maize, file = paste0("/Users/jordyn/Desktop/BCB546_Spring2025/assignments/R-Assignment/chr", i, "_decreasing_maize.txt"), 
              row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
} # Writes to a text file
```

#### same thing but for teosinte data

```{r}
for (i in 1:10) {
  dec_teosinte <- subset(joined_teosinte, joined_teosinte[[3]] == i)
  dec_teosinte[dec_teosinte == "?"] <- "-"  # Replace "?" with "-"
  dec_teosinte <- dec_teosinte[order(dec_teosinte[[2]], decreasing = TRUE), ] # Sorts the data by the second column in decreasing order
  assign(paste0("chr", i, "_decreasing_teosinte"), dec_teosinte) # Saves as a new variable in the environment
  write.table(dec_teosinte, file = paste0("/Users/jordyn/Desktop/BCB546_Spring2025/assignments/R-Assignment/chr", i, "_decreasing_teosinte.txt"), 
              row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
} # Writes the result to a file
```

#### unknown positions for maize, this is looking at rows where the third column has a "?"

```{r}
write.table(subset(joined_maize, joined_maize[[3]] == "?"), 
            file = file.path("/Users/jordyn/Desktop/BCB546_Spring2025/assignments/R-Assignment", "unknown_positions_maize.txt"), 
            row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
```

#### unknown positions for teosinte, this is also looking at rows where the third column has a "?"

```{r}
write.table(subset(joined_teosinte, joined_teosinte[[3]] == "?"), 
            file = file.path("/Users/jordyn/Desktop/BCB546_Spring2025/assignments/R-Assignment", "unknown_positions_teosinte.txt"), 
            row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
```

## Data Visualization: Part II

```{r}
library(ggplot2)
library(dplyr)
```

#### Combine maize and teosinte data with a new "Type" column which is either maize or teosinte

```{r}
maize_type_data <- joined_maize %>% mutate(Type = "Maize")
teosinte_type_data <- joined_teosinte %>% mutate(Type = "Teosinte")
```

#### Merge teosinte and maize datasets

```{r}
combined_data <- bind_rows(maize_type_data, teosinte_type_data)
```

#### This creates the plot that shows the distribution of SNPs across chromosomes in both species

```{r}
ggplot(combined_data, aes(x = as.factor(combined_data$Chromosome), fill = Type)) +
  geom_bar(position = "dodge") +
  labs(title = "SNPs across Chromosomes",
       x = "Chromosome",
       y = "SNP #",
       fill = "Species") +
  theme_minimal() +
  scale_fill_manual(values = c("Maize" = "blue", "Teosinte" = "red"))
```

#### From what I can tell based on the graph it seems as though teosinte and maize have the same number of SNPs in each chromosome
